---
title: "TOG: What is R? Workshop"
date: "Jan 19th 2022"
author: "Divya Anantsri"
output: 
  html_document: 
    keep_md: yes #keep the .md version of this .rmd
    theme: cosmo  #the themes your rmd-to-html document
    highlight: kate #your code highlight theme
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```  

***  

### 0.0 Introduction  

Welcome to the BCCHR Trainee 'Omics Group's `"What is R?"` workshop! 

In this **tutorial**, we are covering: 

1. installing R studio 
2. layout of R studio 
3. introduction to R markdown 
4. setting up - installing packages and loading libraries 
4. exploratory data analysis (EDA) using a sample dataset 


Our **goals** for the end of today are: 

1. for you to feel comfortable navigating R studio and working with R markdown  
2. to give you the ability to self-guide online tutorials or attend future workshops  
  
  
*Please also refer to the accompanying powerpoint slides for a list of useful resources!*    

***  

### 1.0 Getting help in R  

```{r help, eval=FALSE}

help(mean)

?mean

vignette()

```

***  

### 2.0 Installing packages 

```{r install-packages, eval=FALSE}

install.packages("tidyverse")
#this a collection of packages that we will use for data manipulation and plotting 

install.packages("palmerpenguins")
#this package has the sample dataset we will use - Palmerpenguins dataset 
#this dataset is great for learning exploratory data analysis 

```

***  

### 3.0 Load the respective libraries

```{r loading-packages}

library(tidyverse)
library(palmerpenguins)

```

***  

### 4.0 Look at the dataset (penguins)

```{r eda}

#making a new object/vector using the `penguins` dataset that is pre-built into the palmerpenguins package
penguins <- palmerpenguins::penguins

head(penguins)
#view the first few entries in your dataset

tail(penguins)
#view the last few entries

glimpse(penguins)
#view the structure of each variable/vector in the dataset

```

***  

### 5.0 What is the structure of this dataset?

```{r structure-dataset}

str(penguins)
#str is the baseR fucntion of the glimpse fucntion we used above

```

***  

### 6.0 Using ggplot2 for exploratory analysis of the `palmerpenguins` dataset 

Code from https://www.reed.edu/data-at-reed/resources/R/scatterplots.html

ggplot2 syntax includes:  

1. data = dataset being plotted
2. aes = aesthetic mappings - specifiying the variables for the X and Y axes
3. geom = genometric objects - points, lines, bars 

```{r scatterplot}

ggplot(data = penguins, 
       mapping = aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point()

# using the tidyverse `pipe` to select our dataset and the ggplot `+` to add layers and specifications of plotting
# penguins %>% 
#   ggplot(aes(x = bill_length_mm, y = body_mass_g)) +
#   geom_point()

```

Add the visual variable of color to show patterns within each island 

```{r colour-scatterplot}

ggplot(data = penguins, 
       mapping = aes(x = bill_length_mm, y = body_mass_g, color = island)) +
  geom_point()

# using the tidyverse `pipe` to select our dataset and the ggplot `+` to add layers and specifications of plotting
# penguins %>% 
#   ggplot(aes(x = bill_length_mm, y = body_mass_g, colour = island)) +
#   geom_point()

```  

***  

### 7.0 Store your project information in .Rproj   

Whenever you start a project or analysis, it is best practice to make a separate directory and sub-directories containing all of your files. An example would be:  

`project_folder`  
|  
|__ `data` (where you store all your data files; you can also divide the folder further into `raw`, `processed`, `external`)  
|  
|__ `scripts` (all your .RMD, .MD, and HTML/Word?PDF knitted files)  

You can then store all of the directory information in a `.Rproj` file (and you can open that .Rproj file whenever working on a project), by clicking on the blue cube with the R and green plus sign in the option bar above --> Existing Directory --> navigate to your `project_folder`. The top level, i.e. your `project_folder` will be set as your working directory for that specific .Rproj

***  

### 8.0 Knit your file!  

```{r knit, eval=FALSE}

#rmarkdown::render("what_is_R_Jan_2022.Rmd", "html_document")

```  

Easier way to Knit:  

- Click on the Knit symbol dropdown arrow in the option bar above --> Knit Directory --> Project Directory (to ensure that you set the Knit parameter to your .Rproj folder directory)
- Click on the Knit symbol  

OR  

- Ctrl/Cmd + Shift + K  

***  

