---
title: "Introduction to data wrangling in R"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections:  true
    theme: paper
    self_contained: yes
editor_options: 
  chunk_output_type: inline
---

Adapted from Tutorial by: Victor Yuan

Who: Almas Khan, [Trainee Omics Group (TOG)](https://bcchr-trainee-omics-group.github.io/)

What: Introduction to data wrangling in R

Where: BC Children's Hospital Research Institute

When: July 23,2020

# Introduction 

## Who is this tutorial for?

- Those interested using R for data analysis
- Beginners in R, and those curious about `dplyr` and **tidyverse**

## Setup

- Please have the latest version of R and Rstudio installed. 
- Download the [r markdown file](https://github.com/BCCHR-trainee-omics-group/StudyGroup/blob/master/workshops/2020-07-23_data_wrangling_ak/completed_worksheet/Completed_instructor_data_wrangle_workshop.Rmd) for this workshop and open it in Rstudio.

- Install the following R packages now, if you haven't already:

```{r, eval = F}
install.packages(c('gapminder', 'tidyverse'))
```


## Overall Learning outcomes

- Familiarity with tidyverse syntax
- Understanding how to use dplyr to manipulate data in R
- "Piping"

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```



To learn data manipulation in R, some understanding of how to use R is required. A comprehensive lesson on R is out of the scope of this tutorial, but here I go over some of the basics that will be helpful to follow along the data wrangling portion of this workshop.


## Resources

- [R Swirl](https://swirlstats.com/) for interactive lesson on programming with R
- *R for data science* [tibbles chapter](http://r4ds.had.co.nz/tibbles.html) to learn more about tibble/data.frames
- [Learnr package](https://rstudio.github.io/learnr/) another interactive lesson within R on programming


I recommend using **r markdown file (recommended)** for all of your analysis scripts. Think of it as a like a notebook for your code that includes comments and I find it's easier to keep track of your analysis. 

In `.rmd` files, code needs to be in code chunks (insert code chunk shortcut is Windows: `ctrl` `+` `alt` `+` `i`, Mac: `cmd` `+` `alt` `+` `i`) in order to run.


## Error messages (4 min)

As you work with R you will encounter error messages along the way, here are some steps I recommend you follow: 

1. Inspect your code for simple mistakes. Examples: are you missing a bracket or quotation mark? Did you mean to use `=` instead of `==`? Did you spell the name of the function or object wrong?
2. Are using the function correctly? Use the `?` command to pull up the help page. 
3. Make sure your input is correct. Does the function expect a `data.frame` but you gave it a `vector`?
4. Copy and paste the error and the name of the function you are using into **Google**.

# Understand the data in base R: Intro to data types and initial exploration

## Topics

- Basic data types
- `data.frame` and `tibble` objects
- Basic functions for exploring dataframes

Before we begin to manipulate data we should understand what type of data we have:

In R there are multiple types of data and I have listed a couple of basic ones below: 

```{r}
3.14 # This is a numeric 
"This is a character" # A character type 
"3.14" # This is also a character 
TRUE # logical
FALSE # These are logicals
```

These different types of data can be combined into objects like data frames or tibbles:

## Data frames and tibbles (5 min)

`data.frame` objects are by far the most common and useful way to work with your data in R.

A data.frame is basically just a table, it has a certain number of rows, and a certain number of columns.

"trees" is a built-in dataset in R available as a `data.frame`.

```{r}
trees
```

The columns of the trees `data.frame` object are individual `vector` objects. So trees has 3 columns/vectors that are each 31 elements long. The dbl is just a type of numeric class of data. 

### Some basic functions to help understand your `data.frame` objects are:

```{r,error=TRUE}
# number of rows
nrow(trees)
# number of columns
ncol(trees)
# row x columns
dim(trees)
# some basic info on the "structure" of the data.frame
str(trees)
# calculates some summary statistics on each column
summary(trees)
# print first 6 rows
head(trees)
# print last 6 rows
tail(trees)
```

### Tibbles

We are going to use the "gapminder" dataset today, which is stored as a special type of `data.frame`, called a [`tibble`](https://tibble.tidyverse.org/index.html). 

- tibbles have a special printing output
- tibbles never have row names
- any function that works with `data.frame` objects (e.g. `str`, `summary`), will also work with `tibble` objects 
- use functions `as_tibble` and `as.data.frame` to convert between `tibble` and `data.frame`

```{r,warning = FALSE, message = FALSE}
library(gapminder) #load gapminder
```

```{r}
# tibble
gapminder
# data.frame 
head(as.data.frame(gapminder), n = 50)
```

Note that I printed only the first 50 rows of the `data.frame` version of gapminder. By default, printing `tibble` objects will display the top 10 rows to avoid over-printing.

# Let's wrangle:  Intro to tidyverse

Now that we've learned some basics about data types in R. Let's wrangle some data using the tidyverse package. 

## Topics

* Relational/comparison and logical operators
* "Base R" vs tidyverse
* dplyr basic functions


## Resources

Most of this material is inspired from the [STAT 545 `dplyr` lecture](https://stat545guidebook.netlify.com/intro-to-data-wrangling-part-i.html)

Other resources for `dplyr`:

* [Old STAT 545 `dplyr` tutorial](http://stat545.com/block009_dplyr-intro.html)
* [The `dplyr` vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html)

### R Operators (4 min)

Before diving into the tidyverse, let's take a look at a list of built-in operators in R that can help in wrangling our data and using within tidyverse functions. 

See knitted html for pretty tables.

**Arithmetic** operators allow us to carry out mathematical operations:

| Operator | Description |
|------|:---------|
| + | Add |
| - | Subtract |
| * | Multiply |
| / | Divide |
| ^ | Exponent |
| %% | Modulus (remainder from division) |

**Relational** operators allow us to compare values:

| Operator | Description |
|------|:---------|
| < | Less than |
| > | Greater than |
| <= | Less than or equal to |
| >= | Greater than or equal to |
| == | Equal to |
| != | Not equal to |

* Arithmetic and relational operators work on vectors.

**Logical** operators allow us to carry out boolean operations:

| Operator | Description |
|---|:---|
| ! | Not |
| \| | Or (element_wise) |
| & | And (element-wise) |
| \|\| | Or |
| && | And |

## Base-R vs tidyverse

The [tidyverse](https://www.tidyverse.org/) is a collection of packages that were designed under a specific philosophy of making data analysis in R easy, reproducible, and consistent. The tidyverse packages:

* has an easy-to-read syntax
* can be much quicker (to write)  
* are comprehensive in covering a wide variety of applications (e.g. [`ggplot2`](https://ggplot2.tidyverse.org/) for plotting, [`dplyr`](https://dplyr.tidyverse.org/) for data manipulation, ..., etc.)

Most base-r operations can be replaced with tidyverse code. However, in genomics research, this is less true because many genomics analysis workflows rely on [**Bioconductor**](https://www.bioconductor.org/).

[Bioconductor](https://www.bioconductor.org/) is an online R package repository specifically for analyzing and storing genomic data. If you are in genomics, it is highly likely you will use these packages. Bioconductor is basically all based in base-R code, so learning *some* base-R is going to be a requirement for any genomics analysis.


## Tidyverse syntax (2 min)

- Almost all functions from a **tidyverse** package will accept a `tibble` or `data.frame` as the first argument.

```
function(a_tibble, operations_involving_columns)
```

- **tidyverse** functions tend to be [named](https://principles.tidyverse.org/function-names.html) as verbs. (e.g. `select()`, `pivot_longer()`, `str_extract()`)

- More information on the design principles behind tidyverse packages can be found [here](https://principles.tidyverse.org/index.html).

# Basic `dplyr` functions

We're going to be working with the gapminder dataframe we loaded before. Additionally we need to we need to load in the `dplyr` R package. This is a package within the tidyverse that stands for data frame applyer and contains functions to help wrangle data in data.frames or tibbles. 


```{r load packages, warning = FALSE, message = FALSE}
# load your package here:
library(dplyr)
```

We're going to go over these dplyr functions:

1. select 
2. arrange
3. filter 
4. mutate
5. grouped operations in dplyr:
    a. group_by
    b. summarize

### `select()` (8 min)

`select()` allows you to subset to the columns(or variables) that you specify. 

1. Make a data frame containing the columns `year`, `lifeExp`, `country` from the gapminder data, in that order.

```{r}
select(gapminder, year,lifeExp,country)
```


2. Select all variables, from `country` to `lifeExp`.

```{r}
# This will work:
select(gapminder, country, continent, year, lifeExp)
# Better way:
select(gapminder, country:lifeExp)
```


3. Select all variables, except `lifeExp`.

```{r}
select(gapminder, country:year)
```

4. Put `continent` first. Hint: use the `everything()` function.

```{r}
select(gapminder, continent,everything())
```

### `arrange()` (8 min)

1. Order by year.

```{r}
arrange(gapminder, year)
```

2. Order by year, in descending order.

```{r}
arrange(gapminder, desc(year))
```

3. Order by year, then by life expectancy.

```{r}
arrange(gapminder,year, lifeExp)
```

## Piping, `%>%` (8 min)

*Piping* refers to using the `%>%` operator to write nested function calls in a more readable manner.

- Takes an output as the input for the first argument of the next function.

- Think of `%>%` as the word "then"!

**Demonstration:** Here I want to combine `select()` Task 1 with `arrange()` Task 3.

This is how I could do it by *nesting* the two function calls, **without piping**:

```{r nesting functions example, eval = F}
# Nesting function calls can be hard to read
arrange(select(gapminder, year, lifeExp, country), year, lifeExp)
```

Now using **with piping**:

```{r}
# alter the above to include 2 "pipes"
gapminder %>%
  select(year, lifeExp, country) %>%
  arrange(year, lifeExp)
```


## `filter()` (6 min)

Use `filter()` to subset to rows within your data where the condition you specify is TRUE. 

1. Only take data with population greater than 100 million.

```{r}
gapminder %>%
  filter(pop>100000000)
```

2. Your turn: of those rows filtered from step 1., only take data from Asia.

```{r}
gapminder %>%
  filter(pop>100000000 & continent=='Asia')
```

3. Only take data from countries Brazil and China. 

```{r}
gapminder %>%
  filter(pop>100000000,country == "Brazil"| country == "China")
```

## `mutate()` (8 min)

The `mutate()` function _creates_ new columns in the tibble by transforming other variables. Like `select()`, `filter()`, and `arrange()`, the `mutate()` function also takes a tibble as its first argument, and returns a tibble. 

The general syntax is:

```
mutate(tibble, NEW_COLUMN_NAME = CALCULATION)
```

1. Make a new column named `GDP` that equals to multiplying GPD per capita with population.

```{r}
gapminder %>%
  mutate(GDP=gdpPercap*pop)
```

2. Make a new column named `GDP_bill`, that is GDP in billions.

```{r}
gapminder %>%
 mutate(gdpBill = round(gdpPercap*pop/1E9,digits=2))
```

Your turn: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}
gapminder %>%
  mutate(cc = paste(country,continent,sep = ","))
```

# `dplyr`: grouped operations

*Most of this material was taken/modified from the [STAT545 lecture: Intro to data wrangling, Part II](https://stat545guidebook.netlify.com/intro-to-data-wrangling-part-ii.html)*

## Resources

- [STAT545 lecture: Intro to data wrangling, Part II](https://stat545guidebook.netlify.com/intro-to-data-wrangling-part-ii.html)
- [r4ds: transform](http://r4ds.had.co.nz/transform.html)
- [Intro to `dplyr` vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html)

## `summarize()` (3 min)

Like `mutate()`, the `summarize()` function also creates new columns, but the calculations that make the new columns must reduce down to a single number.

For example, let’s compute the mean and standard deviation of life expectancy in the gapminder data set:

```{r}
gapminder %>% 
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))
```

Notice that all other columns were dropped. This is necessary, because there’s no obvious way to compress the other columns down to a single row. This is unlike `mutate(`), which keeps all columns.

As it is, this is hardly useful. But that’s outside of the context of *grouping*, coming up next...

## `group_by()` (15 min)

* `group_by()` allows you to apply functions to separate chunks of your data frame, where the chunks are defined by a specified grouping variable.

Let’s group the gapminder dataset by continent and year:

```{r}
gapminder %>% 
  group_by(continent, year)
```

The only thing different from a regular tibble is the indication of grouping variables above the tibble. This means that the tibble is recognized as having “chunks” defined by unique combinations of continent and year:

- Asia in 1952 is one chunk.
- Asia in 1957 is another chunk.
- Europe in 1952 is another chunk.
- etc…

Notice that the data frame isn’t rearranged by chunk!

Now that the tibble is grouped, operations that you do on a grouped tibble will be done independently within each chunk, as if no other chunks exist.

1. What is the mean and standard deviation of life expectancy for each year for every continent?

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))
```

2. In the gapminder dataset, how many rows are there for each continent? Hint: use the convenience function `dplyr::n()`

```{r}
# solution 1
gapminder %>%
  group_by(continent) %>%
 summarise(n())
# solution 2: use dplyr::count()
gapminder %>%
  count(continent)
```

2. (a) What's the minimum life expectancy for each continent and each year? (b) Arrange by min life expectancy.

```{r}
gapminder %>% 
  group_by(continent,year) %>% 
 summarize(min_life = min(lifeExp)) %>% 
  arrange (min_life)
```

3. Calculate the growth in population since the first year on record _for each country_. Here's another convenience function for you: `dplyr::first()`. 

```{r}
gapminder %>% 
group_by(country) %>% 
arrange(year) %>% 
mutate(rel_growth = pop-first(pop)) 
```

