# Introduction 

## Who is this tutorial for?

- Those interested using R for data analysis
- Beginners in R, and those curious about `dplyr` and **tidyverse**

## Setup

- Please have the latest version of R and Rstudio installed. 
https://cran.r-project.org/


I recommend using **r markdown file (recommended)** for all of your analysis scripts. Think of it as a like a notebook for your code that includes comments and I find it's easier to keep track of your analysis. 


## Overall Learning outcomes

- Familiarity with tidyverse syntax
- Understanding how to use dplyr to manipulate data in R
- "Piping"


## Topics

- Basic data types
- `data.frame` and `tibble` objects
- Basic functions for exploring dataframes


- Install the following R packages now, if you haven't already:

```{r, eval = F}
install.packages(c('gapminder', 'tidyverse'))
```


Before we begin to manipulate data we should understand what type of data we have:

In R there are multiple types of data and I have listed a couple of basic ones below: 

```{r}
3.14 # This is a numeric 
"This is a character" # A character type 
"3.14" # This is also a character 
TRUE # logical
FALSE # These are logicals
```

These different types of data can be combined into objects like data frames or tibbles:

## Data frames and tibbles (5 min)

`data.frame` objects are by far the most common and useful way to work with your data in R.

A data.frame is basically just a table, it has a certain number of rows, and a certain number of columns.

"trees" is a built-in dataset in R available as a `data.frame`.

```{r}
trees
```

The columns of the trees `data.frame` object are individual `vector` objects. So trees has 3 columns/vectors that are each 31 elements long. 

### Some basic functions to help understand your `data.frame` objects are:

```{r,error=TRUE}
# number of rows
nrow(trees)
# number of columns
ncol(trees)
# row x columns
dim(trees)
# some basic info on the "structure" of the data.frame
str(trees)
# calculates some summary statistics on each column
summary(trees)
# print first 6 rows
head(trees)
# print last 6 rows
tail(trees)
```

### Tibbles

Tibbles is special type of data frame.

- tibbles have a special printing output
- tibbles never have row names
- any function that works with `data.frame` objects, will also work with `tibble` objects 
- use functions `as_tibble` and `as.data.frame` to convert between `tibble` and `data.frame`

We are going to use the "gapminder" dataset today.


```{r,warning = FALSE, message = FALSE}
library(gapminder) #load gapminder
```

```{r}
# tibble
gapminder
# data.frame 
head(as.data.frame(gapminder), n = 50)
```


# Let's wrangle:  Intro to tidyverse

Now that we've learned some basics about data types in R. Let's wrangle some data using the tidyverse package. 

## Topics

* Relational/comparison and logical operators
* "Base R" vs tidyverse
* dplyr basic functions


### R Operators (4 min)

Before diving into the tidyverse, let's take a look at a list of built-in operators. 


**Arithmetic** operators allow us to carry out mathematical operations:

| Operator | Description |

| + | Add |
| - | Subtract |
| * | Multiply |
| / | Divide |
| ^ | Exponent |

```{r}
2+2
4-2
2*6
9/3
3^2
```

**Relational** operators allow us to compare values:

| Operator | Description |

| < | Less than |
| > | Greater than |
| <= | Less than or equal to |
| >= | Greater than or equal to |
| == | Equal to |
| != | Not equal to |

```{r}
2<1
2>1
2>=2
2<=2
2==2
2!=3
```


**Logical** operators allow us to carry out boolean operations:

| Operator | Description |

| ! | Not |
| | | Or (element_wise) |
| & | And (element-wise) |


## Tidyverse syntax (2 min)

- Almost all functions from a **tidyverse** package will accept a `tibble` or `data.frame` as the first argument.

The tidyverse is a collection of packages that were designed under a specific philosophy of making data analysis in R easy, reproducible, and consistent. The tidyverse packages:

has an easy-to-read syntax
can be much quicker (to write)
are comprehensive in covering a wide variety of applications (e.g. ggplot2 for plotting, dplyr for data manipulation, ..., etc.)


# Basic `dplyr` functions

We're going to be working with the gapminder dataframe we loaded before. Additionally we need to we need to load the `dplyr` R package. This is a package within the tidyverse that stands for data frame applyer and contains functions to help wrangle data in data.frames or tibbles. 

```{r load packages, warning = FALSE, message = FALSE}
# load your package here:
library(dplyr)
```

We're going to go over these dplyr functions:

1. select 
2. arrange
3. filter 
4. mutate
5. grouped operations in dplyr:
    a. group_by
    b. summarize


### `select()` (8 min)

`select()` allows you to subset to the columns(or variables) that you specify. 

1. Make a data frame containing the columns `year`, `lifeExp`, `country` from the gapminder data, in that order.

```{r}
select(gapminder, year,lifeExp,country)
```


2. Select all variables, from `country` to `lifeExp`.

```{r}
# This will work:
select(gapminder, country, continent, year, lifeExp)
# Better way:
select(gapminder, country:lifeExp)
```


3. Select all variables, except `lifeExp`.

```{r}
select(gapminder, !lifeExp)
```

4. Put `continent` first. Hint: use the `everything()` function.

```{r}
select(gapminder, continent,everything())
```

### `arrange()` (8 min)

1. Order by year.

```{r}
arrange(gapminder, year)
```

2. Order by year, in descending order.

```{r}
arrange(gapminder, desc(year))
```


3. Order by year, then by life expectancy.

```{r}
arrange(gapminder,year, lifeExp)
```

## Piping, `%>%` (8 min)


*Piping* refers to using the `%>%` operator to write nested function. It makes the chunk more readable.

- Takes an output as the input for the first argument of the next function.

- Think of `%>%` as the word "then"!

**Demonstration:** Here I want to combine `select()` Task 1 with `arrange()` Task 3.

This is how I could do it by *nesting* the two function calls, **without piping**:

```{r nesting functions example, eval = F}
# Nesting function calls can be hard to read
arrange(select(gapminder, year, lifeExp, country), year, lifeExp)
```

Now using **with piping**:

```{r}
# alter the function above to include 2 "pipes"
gapminder %>% 
  select( year, lifeExp, country) %>%
  arrange(year, lifeExp)

```


## `filter()` (6 min)

Use `filter()` to subset to rows within your data where the condition you specify is TRUE. 

1. Only take data with population greater than 100 million.

```{r}
gapminder %>%
  filter(pop>100000000)
```

2. Your turn: of those rows filtered from step 1., only take data from Asia.

```{r}
gapminder %>%
  filter(pop>100000000 & continent=='Asia')
```

3. Your turn: Of those rows filtered from step 1. only take data from countries Brazil OR China. 

```{r}
gapminder %>%
  filter(pop>100000000,country == "Brazil"| country == "China")
```

## `mutate()` (8 min)

The `mutate()` function _creates_ new columns in the tibble by transforming other variables.


1. Make a new column named `GDP` that equals to multiplying GPD per capita with population.

```{r}
gapminder %>%
  mutate("GDP" = gdpPercap * pop)
```

2. Make a new column named `GDP_bill`, that is GDP in billions.

```{r}
gapminder %>%
  mutate("GDP_bill" = gdpPercap*pop/1E9)
```

Your turn: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}
gapminder %>%
  mutate("cc" = paste(country,continent, sep = ","))
```


## `summarize()` (3 min)

Like `mutate()`, the `summarize()` function also creates new columns, but the calculations that make the new columns must reduce down to a single number.

For example, let’s compute the mean and standard deviation of life expectancy in the gapminder data set:

```{r}
gapminder %>% 
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))
```

Notice that all other columns were dropped. This is necessary, because there’s no obvious way to compress the other columns down to a single row. This is unlike `mutate(`), which keeps all columns.


## `group_by()` (15 min)

* `group_by()` allows you to apply functions to separate chunks of your data frame, where the chunks are defined by a specified grouping variable.

1. Let’s group the gapminder dataset by continent and year:

```{r}
gapminder %>% 
  group_by(continent, year)
```

- Asia in 1952 is one chunk.
- Asia in 1957 is another chunk.
- Europe in 1952 is another chunk.
- etc…

Now that the tibble is grouped, operations that you do will be done independently within each chunk.

2. What is the mean and standard deviation of life expectancy for each year for every continent?

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
   summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))
```

3. In the gapminder dataset, how many rows are there for each continent? Hint: use the convenience function `dplyr::n()`

```{r}
# solution 1
gapminder %>%
  group_by(continent) %>%
 summarise(n())
# solution 2: use dplyr::count()
gapminder %>%
  count(continent)
```

4. (a) What's the minimum life expectancy for each continent and each year? (b) Arrange by min life expectancy.

```{r}
gapminder %>% 
  group_by(continent,year) %>% 
 summarize(min_life = min(lifeExp)) %>% 
  arrange (min_life)
```

5. Calculate the growth in population since the first year on record _for each country_. Here's another convenience function for you: `dplyr::first()`. 

```{r}
gapminder %>% 
group_by(country) %>% 
arrange(year) %>% 
mutate(rel_growth = pop-first(pop)) 
```

```{r}
gapminder %>% 
group_by(country) %>% 
arrange(year) %>% 
mutate(rel_growth = pop-first(pop)) %>% 
summarise(rel_growth= mean(rel_growth))
```
